---
title: "Merge barcode-variant tables for libraries combined into different experimental pools"
author: "Tyler Starr"
date: "02/20/2024"
output:
  github_document:
    toc: true
    html_preview: false
editor_options: 
  chunk_output_type: inline
---
This notebook takes different pools as defined in the config.yaml file and makes a merged barcode-variant lookup table that pools different libraries according to experimental specifications.

```{r setup, message=FALSE, warning=FALSE, error=FALSE}
#list of packages to install/load
packages = c("yaml","data.table","tidyverse","gridExtra","plotly","withr","htmlwidgets","knitr")
#install any packages not already installed
installed_packages <- packages %in% rownames(installed.packages())
if(any(installed_packages == F)){
  install.packages(packages[!installed_packages],
                   lib=c(paste("/uufs/chpc.utah.edu/common/home/",Sys.getenv("USER"),"/RLibs/",Sys.getenv("R_VERSION"),sep="")),
                   repos=c("http://cran.us.r-project.org"))
}
#load packages
invisible(lapply(packages, library, character.only=T))

knitr::opts_chunk$set(echo = T)
knitr::opts_chunk$set(dev.args = list(png = list(type = "cairo")))

#read in config file
config <- read_yaml("config.yaml")

```
Session info for reproducing environment:
```{r print_sessionInfo}
sessionInfo()
```

Read in tables of per-library barcode-variant lookups, and store the underlying "library" in a new column called "sublibrary"

```{r input_data}
#read in barcode-variant lookup tables
dt_MERS <- data.table(read.csv(file=config$codon_variant_table_file_MERS,stringsAsFactors=F)); dt_MERS[,sublibrary:=library]
dt_PDF2180 <- data.table(read.csv(file=config$codon_variant_table_file_PDF2180,stringsAsFactors=F)); dt_PDF2180[,sublibrary:=library]

dt <- rbind(dt_MERS,dt_PDF2180)

head(dt)
```

Make new tables for each pool (better to do 'by pool' than 'by library', since each pool is unique but any library can in theory be part of multiple pools). Note, this code must be manually updated when new pools are designated in the config.yaml file.

```{r dt_by_pool}

#pool1
dt_pool1 <- dt[sublibrary %in% config$pool1]
dt_pool1[,library:="pool1"]
head(dt_pool1)

#pool2
dt_pool2 <- dt[sublibrary %in% config$pool2]
dt_pool2[,library:="pool2"]
head(dt_pool2)

```

Eliminate barcodes that are repeated between different variants within a single pool.

```{r remove_dup_barcodes}

#pool1
duplicates_pool1 <- dt_pool1[duplicated(dt_pool1,by=c("barcode","library")),.(library,barcode)] #the data.table duplciates function annoyingly only flags the first of each duplicate so doesn't intrinsically allow removal of both of the entries of the duplicate. So, flag what are duplicates, and then remove
dt_pool1[,duplicate:=FALSE]
if(nrow(duplicates_pool1) > 0){
  for(i in 1:nrow(duplicates_pool1)){
    dt_pool1[library==duplicates_pool1[i,library] & barcode==duplicates_pool1[i,barcode],duplicate:=TRUE]
  }
}
dt_pool1 <- dt_pool1[duplicate==FALSE,]; dt_pool1[,duplicate:=NULL]
print(paste("Removed", nrow(duplicates_pool1), "repeated barcodes from pool1"))

#pool2
duplicates_pool2 <- dt_pool2[duplicated(dt_pool2,by=c("barcode","library")),.(library,barcode)] #the data.table duplciates function annoyingly only flags the first of each duplicate so doesn't intrinsically allow removal of both of the entries of the duplicate. So, flag what are duplicates, and then remove
dt_pool2[,duplicate:=FALSE]
if(nrow(duplicates_pool2) > 0){
  for(i in 1:nrow(duplicates_pool2)){
    dt_pool2[library==duplicates_pool2[i,library] & barcode==duplicates_pool2[i,barcode],duplicate:=TRUE]
  }
}
dt_pool2 <- dt_pool2[duplicate==FALSE,]; dt_pool2[,duplicate:=NULL]
print(paste("Removed", nrow(duplicates_pool2), "repeated barcodes from pool2"))

```

Merge the per-pool tables back into one aggregate data table and save.

```{r aggregate_and_save}

dt_final <- rbind(dt_pool1, dt_pool2)

dt_final %>%
  write.csv(file=config$codon_variant_table_file_pools, row.names=F)


```

